---
title: "Activity Monitoring Data"
author: "Daniel Frederico Lins Leite"
date: "20 November 2016"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(ggplot2)
```

## Load and process data

We will analyze the data set "Activity monitoring data"" (amdata) that contains data at 5 minute intervals from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

```{r echo=TRUE}
unzip("activity.zip")
amdata <- read.csv('activity.csv')
completeAmData <- amdata[complete.cases(amdata$steps),]

qtyNA <- sum(is.na(amdata$steps))
pNA <- qtyNA / nrow(amdata)
qtyZero <- sum(completeAmData$steps == 0)
pZero <- qtyZero / nrow(amdata)
```

The data set contains both NA and zero values in the step colums:  
*Number of NA:* `r qtyNA` (`r format(round(pNA,2), nsmall=2)`%)  
*Number of Zero:* `r qtyZero` (`r format(round(pZero,2), nsmall=2)`%)  

To better understand the distribution of steps we will generate two histograms, specially because of the high number of both NA and zero values.

```{r}
completewithzero <- completeAmData
completewithzero$Cat <- rep("With Zero", nrow(completewithzero))

completewithoutzero <- completeAmData[completeAmData$steps > 0,]
completewithoutzero$Cat <- rep("Without Zero", nrow(completewithoutzero))

histData <- merge(completewithzero, completewithoutzero, all = TRUE)
histData$Cat = ordered(histData$Cat, levels = c("With Zero","Without Zero"))

ggplot(histData, aes(x = steps)) +
  geom_histogram( binwidth = 30) +
  facet_grid(Cat ~ ., scales = "free_y") +
  xlab("Steps") +
  ylab("# steps") +
  ggtitle("Steps Distribution")
```

## Steps by day

To better understand the behaviour of the individual, we can also analyze the distribution of the sum of steps by day.

```{R}
stepsByDay <- completeAmData %>%
  group_by(date) %>%
  summarize(steps=sum(steps))

completedMean <- mean(stepsByDay$steps)
completedMedian <- median(stepsByDay$steps)

ggplot(stepsByDay, aes(x = steps)) +
  geom_histogram(binwidth = 2000) +
  ggtitle("Steps by Day") +
  xlab("Steps") +
  ylab("# days")
```

Summary:  
*Mean:* `r format(completedMean, nsmall = 2)`  
*Median:* `r format(completedMedian, nsmall = 2)`

```{R}
meanStepsByInterval <- completeAmData %>%
  group_by(interval) %>%
  summarize(steps=mean(steps))
maximum <- meanStepsByInterval[meanStepsByInterval$steps == max(meanStepsByInterval$steps),]

maximumTick <- sprintf("%02i:%02i", round(maximum$interval/100,0), (maximum$interval %% 100))

ggplot(meanStepsByInterval, aes(x = interval, y = steps)) +
  geom_line() +
  geom_point(aes(x = maximum$interval, y = maximum$steps), colour = 'red') +
  scale_x_discrete(name ="Hour", 
                    breaks=c(0, 500, 1000, 1500, 2000, maximum$interval),
                   limits=c(0, 500, 1000, 1500, 2000, maximum$interval),
                   labels=c("00:00", "05:00", "10:00", "15:00", "20:00", maximumTick)
                   ) +
  ggtitle("Mean Steps per Interval")
```

Summary of Mean Steps per Day:  
Maximum `r format(maximum$steps, nsmall=2)` steps at `r maximum$interval` interval

## Analyzing NA values

The data set contains `r sum(is.na(amdata$steps))` lines with empty value for the steps.

This values will be deduced using the "mean step by interval" above. Every row that contains a NA at the step column will be equal to the mean step count of that interval.  

For this we will use the Coalesce method, that will always choose the first non-NA, non-NULL value in the order specified.

```{R}

coalesce <- function(...) {
  Reduce(function(x, y) {
    i <- which(is.na(x))
    x[i] <- y[i]
    x},
    list(...))
}
```
```{R echo=TRUE}

filledData <- amdata
filledData$steps <- coalesce(filledData$steps,
  as.integer(
    left_join(filledData, meanStepsByInterval, by = 'interval')$steps.y))
```

With this transformation, the data set now contains the following histogram.

```{R}
filledStepsByDay <- filledData %>%
  group_by(date) %>%
  summarize(steps=sum(steps))

ggplot(filledStepsByDay, aes(x = steps)) +
  geom_histogram(binwidth = 2000) +
  ylab("# days") +
  ggtitle("Steps by Day (Deducing NA values)")

filledMean <- mean(filledStepsByDay$steps)
filledMedian <- median(filledStepsByDay$steps)
```

Summary of Steps by Day (Deducing NA values):  
*Mean:* `r format(filledMean, nsmall = 2)`  
*Median:* `r format(filledMedian, nsmall = 2)`

## Comparing the impact

We can compare the impact of the change had comparing the mean and the median statistics before and after the modification:  

*Mean*  
- Before: `r format(completedMean, nsmall = 2)`    
- After: `r format(filledMean, nsmall = 2)`    
*Median*  
- Before: `r format(completedMedian, nsmall = 2)`    
- After: `r format(filledMedian, nsmall = 2)`    

## Analysis on Weekends

To better understand the individual we can also analyze his behaviour comparing the quantity of steps monitored on weekdays and on weekends.

```{R}

filledData$weekday <- weekdays(as.Date(filledData$date))
filledData$weekday <- as.factor(ifelse(filledData$weekday %in% c('Saturday',"Sunday"), "Weekend", "Weekday"))

filledStepsByInterval = filledData %>%
  group_by(interval,weekday) %>%
  summarize(steps=mean(steps))

ggplot(filledStepsByInterval, aes(x = interval, y = steps)) +
  geom_line() +
  ylab("# steps") +
  facet_grid(weekday ~ .) +
  scale_x_discrete(name ="Hour", 
                    breaks=c(0, 500, 1000, 1500, 2000),
                   limits=c(0, 500, 1000, 1500, 2000),
                   labels=c("00:00", "05:00", "10:00", "15:00", "20:00"))
```